---
title: "Visualization functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  out.width  = "100%"
)
library(ReLTER.inatEnrich)
```

## Overview

This article illustrates the visualization functions available in
`ReLTER.inatEnrich` using a small example dataset.

```{r setup-data}
df            <- ReLTER.inatEnrich::occ_eLTER_legal
site_boundary <- ReLTER.inatEnrich::site_boundary
```

## Taxonomic composition

`iconic_taxa()` shows the number of observations and species per
iNaturalist iconic taxon group.

```{r iconic-taxa}
iconic_taxa(df)
```

## Top observed species

`top_n_species()` displays the most observed species coloured by
iconic taxon group.

```{r top-species}
top_n_species(df, n = 10)
```

## Temporal distribution

### By year

```{r obs-year}
obs_per_year(df)
```

### By month

```{r obs-month}
obs_per_month(df)
```

### By hour of the day

```{r obs-hour}
obs_per_hour(df)
```

## Conservation category

`obs_pie_chart()` summarises species by conservation category
following a fixed priority order: Alien (IAS) > Habitats Directive >
Birds Directive > Other.

```{r pie-chart}
obs_pie_chart(df)
```

## Interactive maps

`species_richness_map()` produces an interactive Leaflet map where each
grid cell is coloured by species richness.

```{r species-richness-map, out.width = "100%"}
species_richness_map(
  df            = occ_eLTER_legal,
  site_boundary = site_boundary,
  cell_size     = 0.001
)
```

`create_leaflet_occ_map()` shows each observation as a circle marker
coloured by iconic taxon group, with a detailed popup per observation.

```{r occ-map, out.width = "100%"}
create_leaflet_occ_map(
  occ_enriched  = occ_eLTER_legal,
  site_boundary = site_boundary
)
```
