---
title: "Visualization functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  out.width  = "100%"
)
library(ReLTER.inatEnrich)
```

## Overview

This article illustrates the visualization functions available in
`ReLTER.inatEnrich` using a small example dataset.

```{r mock-data, echo = FALSE}
df <- tibble::tibble(
  name                          = c("Anas platyrhynchos", "Quercus robur",
                                    "Bombus terrestris", "Anas platyrhynchos",
                                    "Vulpes vulpes", "Ailanthus altissima"),
  taxon.iconic_taxon_name       = c("Aves", "Plantae", "Insecta",
                                    "Aves", "Mammalia", "Plantae"),
  observed_on_details.year      = c(2020L, 2021L, 2021L, 2022L, 2022L, 2023L),
  observed_on_details.month     = c(3L, 6L, 6L, 3L, 11L, 5L),
  observed_on_details.hour      = c(8L, 14L, 14L, 9L, 17L, 10L),
  taxon_geoprivacy              = c("open", "open", "obscured",
                                    "open", "open", "open"),
  quality_grade                 = rep("research", 6),
  observed_on                   = as.Date(c("2020-03-01", "2021-06-15",
                                            "2021-06-20", "2022-03-10",
                                            "2022-11-05", "2023-05-01")),
  public_positional_accuracy    = c(10, 20, 15, 10, 30, 25),
  user.login                    = c("user_a", "user_b", "user_a",
                                    "user_c", "user_b", "user_a"),
  uri                           = paste0("https://www.inaturalist.org/observations/", 1:6),
  taxon.id                      = c(1L, 2L, 3L, 1L, 4L, 5L),
  taxon.preferred_common_name   = c("Mallard", "English oak", "Buff-tailed bumblebee",
                                    "Mallard", "Red fox", "Tree of heaven"),
  taxon.default_photo.square_url = rep("https://example.com/photo.jpg", 6),
  directive  = c("EU Birds Directive", NA, NA, "EU Birds Directive",
                 NA, "EU Habitats Directive"),
  annex      = c("I", NA, NA, "I", NA, "II"),
  establishmentMeans = list(
    tibble::tibble(nativeness = "native",     authority = "GBIF"),
    tibble::tibble(nativeness = "native",     authority = "GBIF"),
    tibble::tibble(nativeness = "native",     authority = "GBIF"),
    tibble::tibble(nativeness = "native",     authority = "GBIF"),
    tibble::tibble(nativeness = "native",     authority = "GBIF"),
    tibble::tibble(nativeness = "introduced", authority = "GBIF")
  ),
  status_IUCN = replicate(6, tibble::tibble(
    status    = "LC",
    authority = "IUCN",
    name      = "",
    url       = "https://www.iucnredlist.org"
  ), simplify = FALSE),
  lon = c(11.1, 11.2, 11.3, 11.1, 11.4, 11.5),
  lat = c(45.1, 45.2, 45.3, 45.1, 45.4, 45.5)
) |>
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

## Taxonomic composition

`iconic_taxa()` shows the number of observations and species per
iNaturalist iconic taxon group.

```{r iconic-taxa}
iconic_taxa(df)
```

## Top observed species

`top_n_species()` displays the most observed species coloured by
iconic taxon group.

```{r top-species}
top_n_species(df, n = 10)
```

## Temporal distribution

### By year

```{r obs-year}
obs_per_year(df)
```

### By month

```{r obs-month}
obs_per_month(df)
```

### By hour of the day

```{r obs-hour}
obs_per_hour(df)
```

## Conservation category

`obs_pie_chart()` summarises species by conservation category
following a fixed priority order: Alien (IAS) > Habitats Directive >
Birds Directive > Other.

```{r pie-chart}
obs_pie_chart(df)
```

## Interactive maps

The interactive map outputs (species richness grid and occurrence map)
are best explored directly:

- [Species richness map](https://oggioniale.github.io/ReLTER.inatEnrich/figures/map_example.html)
