---
title: "ReLTER.inatEnrich — Technical Reference"
description: >
  A technical deep-dive into the ReLTER.inatEnrich enrichment pipeline:
  API calls, nested list-column structures, and data access patterns.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ReLTER.inatEnrich — Technical Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Overview

This article is a technical complement to the
[Getting Started vignette](../articles/ReLTER_inatEnrich.html).
It documents the internal data structures produced at each step of the
enrichment pipeline, the API endpoints queried, and the recommended
patterns for accessing nested list-columns in downstream analyses.

---

## Package architecture

`ReLTER.inatEnrich` is structured around four public functions that operate
sequentially on an `sf` tibble of iNaturalist occurrences:

| Function | Data source | Output column(s) |
|---|---|---|
| `add_iucn_to_occ()` | iNaturalist API `/v1/taxa/{id}` | `status_IUCN` (list-column) |
| `add_nativeness_to_occ()` | iNaturalist API `/v1/taxa/{id}` | `establishmentMeans` (list-column) |
| `add_eunis_legal_to_occ()` | EUNIS species database | `directive`, `Annex` (flat columns) |
| `create_leaflet_occ_map()` | — | Leaflet map object |

All API calls use `httr2` with retry logic (`max_tries = 3`,
`max_seconds = 120`) and return a safe empty tibble on failure, so the
pipeline never interrupts on a single taxon error.

---

## Nested list-column structures

Two of the four enrichment functions produce **list-columns** — columns
where each cell contains a tibble rather than a scalar value. This design
allows multiple records per taxon (e.g. one IUCN assessment per geographic
scope) to coexist in a single row of the occurrence tibble without
information loss.

### `establishmentMeans`

Produced by `add_nativeness_to_occ()`. Each cell contains a **1 × 2
tibble** with the nativeness status and the authority that recorded it,
filtered to the specified country.

```
# str(occ$establishmentMeans[[1]])
tibble [1 × 2] (S3: tbl_df/tbl/data.frame)
 $ nativeness: chr "native"
 $ authority : chr "Italy Check List"
```

**Accessing the values:**

```{r access-em}
library(dplyr)
library(purrr)

# Extract nativeness for all rows as a flat character vector
occ |>
  mutate(
    nativeness = map_chr(establishmentMeans, ~ .x$nativeness[[1]] %||% NA_character_),
    em_authority = map_chr(establishmentMeans, ~ .x$authority[[1]]  %||% NA_character_)
  )

# Unnest to a flat tibble (duplicates rows where needed)
occ |>
  tidyr::unnest(establishmentMeans)
```

**When is it `NA`?**

`nativeness` and `authority` are both `NA_character_` when:

- the taxon has no listed taxa records in iNaturalist, or
- none of the listed taxa records match the specified `country`, or
- the API call failed after the maximum number of retries.

---

### `status_IUCN`

Produced by `add_iucn_to_occ()`. Each cell contains an **N × 4 tibble**,
with one row per geographic scope for which an IUCN assessment exists.
The number of rows varies by taxon.

```
# str(occ$status_IUCN[[1]])
tibble [3 × 4] (S3: tbl_df/tbl/data.frame)
 $ status   : chr [1:3] "LC" "LC" "LC"
 $ authority: chr [1:3] "IUCN Red List" "IUCN Red List" "IUCN Red List"
 $ name     : chr [1:3] NA "Europe" "Mediterranean"
 $ url      : chr [1:3] "https://www.iucnredlist.org/..."
                        "https://www.iucnredlist.org/..."
                        "https://www.iucnredlist.org/..."
```

The `name` field identifies the geographic scope of the assessment:

- `NA` — global assessment
- `"Europe"`, `"Mediterranean"`, etc. — regional assessments

**Accessing the values:**

```{r access-iucn}
# Extract only the global status (name is NA)
occ |>
  mutate(
    iucn_global = map_chr(status_IUCN, function(tbl) {
      global <- tbl |> filter(is.na(name))
      if (nrow(global) == 0) NA_character_ else global$status[[1]]
    })
  )

# Extract all scopes as a flat tibble (one row per taxon × scope)
occ |>
  select(taxon.id, name, status_IUCN) |>
  tidyr::unnest(status_IUCN) |>
  rename(
    scientific_name = name,
    iucn_scope      = name1,   # name column from the nested tibble
    iucn_status     = status
  )

# Get all distinct statuses for a single taxon
occ$status_IUCN[[1]] |>
  mutate(scope = coalesce(name, "Global")) |>
  select(scope, status, url)
```

**IUCN status codes reference:**

| Code | Meaning |
|---|---|
| `EX` | Extinct |
| `EW` | Extinct in the Wild |
| `CR` | Critically Endangered |
| `EN` | Endangered |
| `VU` | Vulnerable |
| `NT` | Near Threatened |
| `LC` | Least Concern |
| `DD` | Data Deficient |
| `NE` | Not Evaluated |

---

### `directive` and `Annex` (flat columns)

Produced by `add_eunis_legal_to_occ()`. Unlike the IUCN and nativeness
data, EUNIS legal information is stored as **flat character columns**. When
a taxon is covered by more than one directive, the occurrence tibble
contains **multiple rows for the same observation** — one per
directive/annex combination.

```
# Example: a taxon covered by both directives
# A tibble: 2 × (n + 2)
#   taxon.id  ...  directive                              Annex
#   <int>          <chr>                                  <chr>
# 1 48484          EU Birds Directive (2009/147/EC)       Annex I
# 2 48484          EU Habitats Directive (92/43/EEC)      Annex IV
```

This means that **row count in `occ_eLTER_legal` may exceed row count in
`occ_eLTER_nativeness`** for taxa with multiple directive entries. Keep
this in mind when computing summaries:

```{r eunis-dedup}
# Count unique observations (not rows) correctly
n_obs <- occ_eLTER_legal |>
  dplyr::distinct(taxon.id, observed_on, user.login) |>
  nrow()

# Summarise directives per taxon without duplicating observations
directives_per_taxon <- occ_eLTER_legal |>
  dplyr::select(taxon.id, directive, Annex) |>
  dplyr::distinct() |>
  dplyr::group_by(taxon.id) |>
  dplyr::summarise(
    n_directives = sum(!is.na(directive)),
    directives   = paste(na.omit(directive), collapse = "; "),
    annexes      = paste(na.omit(Annex),     collapse = "; "),
    .groups      = "drop"
  )
```

---

## API details

### iNaturalist API

Both `add_iucn_to_occ()` and `add_nativeness_to_occ()` query the
iNaturalist v1 taxa endpoint:

```
GET https://api.inaturalist.org/v1/taxa/{taxon_id}
```

The relevant fields extracted from the response are:

- `results[[1]]$conservation_statuses` → IUCN status, authority, URL
- `results[[1]]$listed_taxa` → establishment means, place name, list title

Both functions iterate over **unique taxon IDs only**, so each taxon is
queried exactly once regardless of how many observations exist for it in
the dataset.

### EUNIS species database

`add_eunis_legal_to_occ()` queries the EUNIS API using the taxon's
scientific name to retrieve legal coverage under EU directives. Results
are matched back to the occurrence tibble via `taxon.id`.

---

## Error handling and NA semantics

All three enrichment functions use the same defensive pattern:

```{r error-pattern}
# Conceptual structure shared by all enrichment functions
result <- tryCatch(
  query_external_api(taxon.id),
  error = function(e) {
    warning("Error for taxon.id: ", taxon.id, " — ", conditionMessage(e))
    return(empty_fallback_tibble)   # never returns NULL
  }
)
```

This guarantees that:

1. The pipeline always completes even if individual API calls fail
2. Failed taxa are identifiable by `NA` values in the enrichment columns
3. A `warning()` is emitted for each failure, visible in the R console

To inspect which taxa failed after running the pipeline:

```{r inspect-failures}
# Taxa with no IUCN data
occ |>
  filter(map_lgl(status_IUCN, ~ all(is.na(.x$status))))

# Taxa with no nativeness data
occ |>
  filter(map_lgl(establishmentMeans, ~ is.na(.x$nativeness[[1]])))

# Taxa with no EUNIS legal data
occ |>
  filter(is.na(directive)) |>
  distinct(taxon.id, name)
```

---

## Performance considerations

For large occurrence datasets (> 1 000 unique taxa), the enrichment
functions can take several minutes due to the sequential nature of the API
calls and the retry logic. A few practical tips:

- **Cache intermediate results** between steps using `saveRDS()` /
  `readRDS()` to avoid re-querying on re-runs:

```{r caching}
# Save after the slowest step
saveRDS(occ_eLTER_IUCN,      "cache/occ_eLTER_IUCN.rds")
saveRDS(occ_eLTER_nativeness, "cache/occ_eLTER_nativeness.rds")
saveRDS(occ_eLTER_legal,      "cache/occ_eLTER_legal.rds")

# Reload
occ_eLTER_legal <- readRDS("cache/occ_eLTER_legal.rds")
```

- **Filter before enriching**: apply `sf::st_intersection()` before
  running the enrichment functions so that only taxa actually within the
  site boundary are queried.
- **Rate limits**: the iNaturalist API does not require authentication for
  read-only taxon queries, but it does apply rate limiting. The built-in
  retry logic (`max_tries = 3`, `max_seconds = 120`) handles transient
  failures gracefully.

---

## Session info

```{r session-info, eval = TRUE}
sessionInfo()
```
